### 2018-07-05 paper/3-abstractions.tex
S3 updates

\begin{figure}
\vspace{-1mm}
\begin{minted}[fontsize=\small]{haskell}
-- An abstract store containing a key/value map and persistent build information
data Store i k v -- i = info, k = key, v = value
initialise :: i -> (k -> v) -> Store i k v
getInfo    :: Store i k v -> i
putInfo    :: i -> Store i k v -> Store i k v
getValue   :: k -> Store i k v -> v
putValue   :: Eq k => k -> v -> Store i k v -> Store i k v
getHash    :: Hashable v => k -> Store i k v -> Hash v
hash       :: Hashable v => v -> Hash v -- provides a fast equality check
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Build tasks (see §3.2)
newtype Task  c k v = Task { run :: @\std{forall}@ f. c f => (k -> f v) -> f v }
type    Tasks c k v = k -> Maybe (Task c k v)
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Build system (see §3.3)
type Build c i k v = Tasks c k v -> k -> Store i k v -> Store i k v
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Build system components: a scheduler and a rebuilder (see §5)
type Scheduler c i ir k v = Rebuilder c ir k v -> Build c i k v
type Rebuilder c   ir k v = k -> v -> Task c k v -> Task (MonadState ir) k v
\end{minted}
\vspace{-3mm}
\caption{Type signatures of key build systems abstractions.}\label{fig-types}
\vspace{-3mm}
\end{figure}

@@

\noindent
Here \hs{c} stands for \emph{constraint}, such as \hs{Applicative} or \hs{Monad}.
A \hs{Task} describes a single build task, while \hs{Tasks} associates a
\hs{Task} to every non-input key; input keys are associated with
\hs{Nothing}. The highly-abstracted type\footnote{Readers
familiar with \emph{lenses} or \emph{profunctor optics} might recognise a
familiar pattern. We discuss this in~\S\ref{sec-related-optics}.} \hs{Task} is
best introduced by an example. Consider this \Excel spreadsheet:

@@

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- Applicative functors
pure  :: Applicative f => a -> f a
(<$>) :: Applicative f =>   (a -> b) -> f a -> f b -- Left-associative
(<*>) :: Applicative f => f (a -> b) -> f a -> f b -- Left-associative
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Standard State monad from Control.Monad.State
data State s a
instance Monad (State s)
get       :: State s s
gets      :: (s -> a) -> State s a
put       :: s -> State s ()
modify    :: (s -> s) -> State s ()
runState  :: State s a -> s -> (a, s)
execState :: State s a -> s -> s
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Standard types from Data.Functor.Identity and Data.Functor.Const
newtype Identity a = Identity { runIdentity :: a }
newtype Const m a  = Const    { getConst    :: m }
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
instance Functor (Const m) where
    fmap _ (Const m) = Const m
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
instance Monoid m => Applicative (Const m) where
    pure _              = Const mempty   -- mempty is the identity of the monoid m
    Const x <*> Const y = Const (x <> y) -- <> is the binary operation of the monoid m
\end{minted}
\vspace{-3mm}
\caption{Standard library definitions.}\label{fig-stdlib}
\vspace{-4mm}
\end{figure}

@@

\begin{minted}[xleftmargin=5pt]{haskell}
compute :: Task Monad k v -> Store i k v -> v
compute task store = runIdentity $ run task (\k -> Identity (getValue k store))
\end{minted}

\noindent
Here we do not need any effects in the \hs{fetch} callback to \hs{task}, so
we can use the standard Haskell \hs{Identity} monad (Fig.~\ref{fig-stdlib}).
The use of \hs{Identity} just fixes the `impedance mismatch' between the
function \hs{getValue}, which returns a pure value~\hs{v}, and the \hs{fetch}
argument of the \hs{task}, which must return an \hs{f}~\hs{v} for some \hs{f}.
To fix the mismatch, we wrap the result of \hs{getValue} in the \hs{Identity}
monad: the function
\hs{\@@k}~\hs{->}~\hs{Identity}~\hs{(@@getValue}~\hs{k}~\hs{store)} has the type
\hs{k}~\hs{->}~\hs{Identity}~\hs{v}, and can now be passed to a \hs{task}. The
result comes as \hs{Identity}~\hs{v}, which we unwrap by \hs{runIdentity}.

@@

\noindent
Here \hs{Const} is a standard Haskell type defined in Fig.~\ref{fig-stdlib}. We
instantiate \hs{f} to \hs{Const}~\hs{[@@k]}. So a value of type \hs{f}~\hs{v},
or in this case \hs{Const}~\hs{[@@k]}~\hs{v}, contains no value \hs{v}, but does
contain a list of keys of type \hs{[@@k]} which we use to record dependencies.
The \hs{fetch} callback that we pass to \hs{task} records a single dependency;
and the standard definition of \hs{Applicative} for \hs{Const} (which we give
in Fig.~\ref{fig-stdlib}) combines the dependencies from different parts of the
task. Running the task with \hs{f}~=~\hs{Const}~\hs{[@@k]} will thus
accumulate a list of the task's dependencies -- and that is just what
\hs{dependencies} does:
\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
@\ghci@ dependencies $ fromJust $ sprsh1 "A1"
[@@]
\end{minted}
\begin{minted}[xleftmargin=10pt]{haskell}
@\ghci@ dependencies $ fromJust $ sprsh1 "B1"
["A1", "A2"]
\end{minted}
\vspace{1mm}

@@

\noindent
This implementation uses the standard Haskell \hs{WriterT} \emph{monad
transformer}~\cite{liang1995monad}, which allows us to record additional
information -- a list of key/value pairs \hs{[(@@k,}~\hs{v)]} -- when computing
a task in an arbitrary monad~\hs{m}. We substitute the given \hs{fetch} with a
\hs{trackingFetch} that, in addition to fetching a value, tracks the
corresponding key/value pair. The \hs{task} returns the value of type
\hs{WriterT}~\hs{[(@@k,}~\hs{v)]}~\hs{m}~\hs{v}, which we unwrap by
\hs{runWriterT}. We will use \hs{track} when implementing monadic build systems
with dynamic dependencies, see~\S\ref{sec-implementation-shake}. Here we show an
example of \hs{track}ing monadic tasks when \hs{m}~\hs{=}~\hs{IO}.

>>>>

\begin{figure}
\vspace{-1mm}
\begin{minted}[fontsize=\small]{haskell}
-- An abstract store containing a key/value map and persistent build information
data Store i k v -- i = info, k = key, v = value
initialise :: i -> (k -> v) -> Store i k v
getInfo    :: Store i k v -> i
putInfo    :: i -> Store i k v -> Store i k v
getValue   :: k -> Store i k v -> v
putValue   :: Eq k => k -> v -> Store i k v -> Store i k v
data Hash v
getHash    :: Hashable v => k -> Store i k v -> Hash v
hash       :: Hashable v => v -> Hash v -- provides a fast equality check
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Build tasks (see §3.2)
newtype Task  c k v = Task { run :: @\std{forall}@ f. c f => (k -> f v) -> f v }
type    Tasks c k v = k -> Maybe (Task c k v)
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Build system (see §3.3)
type Build c i k v = Tasks c k v -> k -> Store i k v -> Store i k v
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Build system components: a scheduler and a rebuilder (see §5)
type Scheduler c i ir k v = Rebuilder c ir k v -> Build c i k v
type Rebuilder c   ir k v = k -> v -> Task c k v -> Task (MonadState ir) k v
\end{minted}
\vspace{-3mm}
\caption{Type signatures of key build systems abstractions.}\label{fig-types}
\vspace{-3mm}
\end{figure}

@@

\noindent
Here \hs{c} stands for \emph{constraint}, such as \hs{Applicative} or \hs{Monad}.
A \hs{Task} describes a single build task, while \hs{Tasks} associates a
\hs{Task} to every non-input key; input keys are associated with
\hs{Nothing}. The highly-abstracted type\footnote{Readers
familiar with \emph{lenses} or \emph{profunctor optics} might recognise a
familiar pattern. We discuss this in~\S\ref{sec-related-optics}.} \hs{Task}
describes how to build a value when given a way to build its dependencies, and is
best explained by an example. Consider this \Excel spreadsheet:

@@

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- Applicative functors
pure  :: Applicative f => a -> f a
(<$>) :: Functor     f =>   (a -> b) -> f a -> f b -- Left-associative
(<*>) :: Applicative f => f (a -> b) -> f a -> f b -- Left-associative
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Standard State monad from Control.Monad.State
data State s a
instance Monad (State s)
get       :: State s s
gets      :: (s -> a) -> State s a
put       :: s -> State s ()
modify    :: (s -> s) -> State s ()
runState  :: State s a -> s -> (a, s)
execState :: State s a -> s -> s
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Standard types from Data.Functor.Identity and Data.Functor.Const
newtype Identity a = Identity { runIdentity :: a }
newtype Const m a  = Const    { getConst    :: m }
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
instance Functor (Const m) where
    fmap _ (Const m) = Const m
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
instance Monoid m => Applicative (Const m) where
    pure _              = Const mempty   -- mempty is the identity of the monoid m
    Const x <*> Const y = Const (x <> y) -- <> is the binary operation of the monoid m
\end{minted}
\vspace{-3mm}
\caption{Standard library definitions.}\label{fig-stdlib}
\vspace{-4mm}
\end{figure}

@@

\begin{minted}[xleftmargin=5pt]{haskell}
compute :: Task Monad k v -> Store i k v -> v
compute task store = runIdentity $ run task (\k -> return (getValue k store))
\end{minted}

\noindent
Here we do not need any effects in the \hs{fetch} callback to \hs{task}, so
we can use the standard Haskell \hs{Identity} monad (Fig.~\ref{fig-stdlib}).
The use of \hs{Identity} just fixes the `impedance mismatch' between the
function \hs{getValue}, which returns a pure value~\hs{v}, and the \hs{fetch}
argument of the \hs{task}, which must return an \hs{f}~\hs{v} for some \hs{f}.
To fix the mismatch, we unwrap using the function \hs{runIdentity} of type \hs{Identity}~\hs{a}~\hs{->}~\hs{a},
which both forces the monad to \hs{Identity} and strips it away.

@@

\noindent
Here \hs{Const} is a standard Haskell type defined in Fig.~\ref{fig-stdlib}. We
instantiate \hs{f} to \hs{Const}~\hs{[@@k]}. So a value of type \hs{f}~\hs{v},
or in this case \hs{Const}~\hs{[@@k]}~\hs{v}, contains no value \hs{v}, but does
contain a list of keys of type \hs{[@@k]} which we use to record dependencies.
The \hs{fetch} callback that we pass to \hs{task} records a single dependency;
and the standard definition of \hs{Applicative} for \hs{Const} (which we give
in Fig.~\ref{fig-stdlib}) combines the dependencies from different parts of the
task. Running the task with \hs{f}~=~\hs{Const}~\hs{[@@k]} will thus
accumulate a list of the task's dependencies -- and that is what
\hs{dependencies} does:
\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
@\ghci@ dependencies $ fromJust $ sprsh1 "A1"
[@@]
\end{minted}
\begin{minted}[xleftmargin=10pt]{haskell}
@\ghci@ dependencies $ fromJust $ sprsh1 "B1"
["A1", "A2"]
\end{minted}
\vspace{1mm}

@@

\noindent
This implementation uses the standard Haskell \hs{WriterT} \emph{monad
transformer}~\cite{liang1995monad}, which allows us to record additional
information -- a list of key/value pairs \hs{[(@@k,}~\hs{v)]} -- when computing
a task in an arbitrary monad~\hs{m}. We substitute the given \hs{fetch} with a
\hs{trackingFetch} that, in addition to fetching a value, tracks the
corresponding key/value pair. The \hs{task} returns a value of type
\hs{WriterT}~\hs{[(@@k,}~\hs{v)]}~\hs{m}~\hs{v}, which we unwrap with
\hs{runWriterT}. We will use \hs{track} when implementing monadic build systems
with dynamic dependencies, see~\S\ref{sec-implementation-shake}. Here we show an
example of \hs{track}ing monadic tasks when \hs{m}~\hs{=}~\hs{IO}.
